Specification of Advanced Types Translation

### 2014-03-08 DOCUMENTATION/AdvancedTypes.tex
Integrating the AdvancedTypes.tex and specfication.tex

\documentclass[10pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsfonts}
\usepackage{amssymb}

\usepackage[fleqn]{amsmath}
\usepackage{xspace,framed,here,keystroke}
\usepackage{graphicx}
\usepackage[colorlinks=false]{hyperref}

\usepackage{minted}   %Used to syntax highlight
\definecolor{bg}{rgb}{0.95,0.95,0.95}
\newminted{llvm}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=bg}
\newminted{pascal}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=bg}
\newminted{c}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=bg}

\newcommand{\trad}[2]{\ensuremath{\lVert \textsf{#1} \rVert^{\textit{#2}}}}
\newcommand{\nl}[0]{\ensuremath{\downarrow}}
\newcommand{\mty}[0]{\texttt{""}}
\DeclareMathOperator{\conc}{\diamond}
\DeclareMathOperator{\isdef}{\equiv}
\DeclareMathOperator{\dom}{\mbox{dom}}
\DeclareMathOperator{\lbl}{\mathcal{L}()}
\DeclareMathOperator{\variable}{\mathcal{V}()}
\newcommand{\llvm}[1]{\texttt{#1}}
\newcommand{\B}[1]{\textsf{#1}}
\newcommand{\lalt}[0]{$\langle$\xspace}
\newcommand{\ralt}[0]{$\rangle$\xspace}
\newcommand{\alt}[0]{$\mid\,$}
\newcommand{\ListOf}[1]{$\mbox{#1}^+$}
\newcommand{\nt}[1]{{\normalfont\textit{#1}}}
\newcommand{\Dict}[0]{\mathbb{D}}
\newcommand{\Text}[0]{\mathbb{T}}
\newcommand{\IF}[0]{\textbf{ if }}
\newcommand{\ELSIF}[0]{\textbf{ else if }}
\newcommand{\ELSE}[0]{\textbf{ else }}
\newcommand{\THEN}[0]{\textbf{ then }}
\newcommand{\LET}[0]{\textbf{ let }}
\newcommand{\IN}[0]{\textbf{ in }}
\newcommand{\AND}[0]{\textbf{ and }}
\newcommand{\PH}[1]{\framebox{$#1$}}
\newcommand{\sep}[0]{\otimes}
\newcommand{\intf}[0]{\ensuremath{\mathbb{I}}}
\newcommand{\Global}[0]{\ensuremath{\sf\Gamma}}
\newcommand{\local}[0]{\ensuremath{\sf\lambda}}
\newcommand{\opmap}[0]{\ensuremath{\sf\Omega}}
\newcommand{\developed}[0]{\ensuremath{\textsf{developed}}}
\newcommand{\stateless}[0]{\ensuremath{\textsf{stateless}}}
\newcommand{\importedmodules}[0]{\ensuremath{\textsf{i-mod}}}
\newcommand{\trimportedmodules}[0]{\ensuremath{\textsf{i-mod$\mathsf{\ast}$}}}
\newcommand{\importedinstances}[0]{\ensuremath{\textsf{i-inst}}}
\newcommand{\trimportedinstances}[0]{\ensuremath{\textsf{i-inst$\mathsf{\ast}$}}}
\newcommand{\idx}[0]{\ensuremath{\sf\Pi}}
\newcommand{\state}[0]{\ensuremath{\sf\Theta}}
\newcommand{\self}[0]{\ensuremath{\sf\Sigma}}
\newcommand{\init}[0]{\ensuremath{\mathsf{I}}}
\newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}
\newcommand{\SBPos}[0]{\ensuremath{\sf\varphi}}
\newcommand{\SBType}[0]{\ensuremath{\sf\tau}}

\newcommand{\DD}[1]{\marginpar{\tiny{DD: #1}}}
\newcommand{\VGM}[1]{\marginpar{\tiny{VGM: #1}}}

\author{David Déharbe, Valério Medeiros Jr.}
\title{Specification of Advanced Types Translation}

\begin{document}

\usemintedstyle{friendly}

\section{The abstract syntax tree with support for advanced types}

The supported advanced types are array and structures. The nodes Array and 
Structure can be used to declare a variable, or  declare an element, or to 
select an element in the array or the structure.

\begin{figure}[H]
  \begin{center}
    {\scriptsize
      \frame{
    \begin{tabular}[t]{ccc}
      \begin{tabular}[t]{rcl}
        \hline
        \multicolumn{3}{|c|}{Machine: \B{Machine}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{const} & : & seq \B{Cons} \\
        \B{var} & : & seq \B{Vari} \\
        \B{init} & : & seq \B{Inst} \\
        \B{op} & : & seq \B{Oper} \\
        \B{impl} & : & opt \B{Impl} \\
        \hline
        \multicolumn{3}{|c|}{Implementation: \B{Impl}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{impo} & : & seq \B{Import} \\
        \B{const} & : & seq \B{Cons} \\
        \B{var} & : & seq \B{Vari} \\
        \B{init} & : & seq \B{Inst} \\
        \B{op} & : & seq \B{Oper} \\
        \hline
        \multicolumn{3}{|c|}{Imports: \B{Import}} \\
        \hline
        \B{mach} & : & \B{Machine} \\
        \B{pre} & : & opt \B{Name}
        \\
        \hline
        \multicolumn{3}{|c|}{Constant: \B{Cons}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{type} & : & \B{Type} \\
        \B{val} & : & \B{Value} \\
        \hline
        \multicolumn{3}{|c|}{Variable: \B{Vari}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{type} & : & \B{Type} \\
        \B{scope} & : & \B{Impl} \alt \B{Oper}\\
	\hline
        \multicolumn{3}{|c|}{Operation: \B{Oper}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{inp} & : & seq \B{Vari} \\
        \B{out} & : & seq \B{Vari} \\
        \B{body} & : & \B{Inst}
        \\
       \hline
        \multicolumn{3}{|c|}{Instruction: \B{Inst}} \\
        \hline
        \multicolumn{3}{c}{\B{Blk} \alt \B{VarD} \alt \B{If} \alt \B{BEq} \alt } \\
        \multicolumn{3}{c}{\B{Call} \alt \B{While} \alt \B{Case} \alt \B{Skip}} \\
        \hline
        \multicolumn{3}{|c|}{Block: \B{Blk}} \\
        \hline
        \B{body} & : & seq \B{Inst} \\
        \hline
        \multicolumn{3}{|c|}{Variable declaration: \B{VarD}} \\
        \hline
        \B{vars} & : & seq \B{Name} \\
        \B{body} & : & seq \B{Inst} \\
\end{tabular}
\begin{tabular}[t]{rcl}   
  	\hline
        \multicolumn{3}{|c|}{If: \B{If}} \\
        \hline
        \B{branches} & : & seq \B{IfBr} \\
        \hline
        \multicolumn{3}{|c|}{Becomes equal: \B{Beq}} \\
        \hline
        \B{lhs} & : & \B{LValue} \\
        \B{rhs} & : & \B{Exp} \\
	\hline
	\multicolumn{3}{|c|}{Left value: \B{LValue}} \\	
	\hline
	\multicolumn{3}{c}{ \B{Vari} \alt \B{ArrayItem} \alt \B{StructItem} }\\
        \hline
        \multicolumn{3}{|c|}{Operation call: \B{Call}} \\
        \hline
        \B{op} & : & \B{Oper} \\
        \B{inp} & : & seq \B{Exp} \\
        \B{out} & : & seq \B{Name} \\
        \B{inst} & : & opt \B{Impo} \\
        \hline
        \multicolumn{3}{|c|}{Case conditional: \B{Case}} \\
        \hline
        \B{expr} & : & \B{Exp} \\
        \B{branches} & : & seq \B{CaseBr}\\
	\hline
        \multicolumn{3}{|c|}{Case branch: \B{CaseBr}} \\
        \hline
        \B{val} & : & seq \B{Value} \\
        \B{body} & : & \B{Inst}
        \\
        \hline
        \multicolumn{3}{|c|}{While loop: \B{While}} \\
        \hline
        \B{cond} & : & \B{Pred} \\
        \B{body} & : & \B{Inst} \\
        \hline
        \multicolumn{3}{|c|}{If branch: \B{IfBr}} \\
        \hline
        \B{cond} & : & opt \B{Pred} \\
        \B{body} & : & \B{Inst} \\
        \hline
        \multicolumn{3}{|c|}{Expression: \B{Exp}} \\
        \hline
        \multicolumn{3}{c}{\B{Lit}\alt\B{Term}\alt\B{Pred}\alt\B{Cons}\alt\B{Vari}}\\
	\multicolumn{3}{c}{\B{Record}\alt\B{Array}\alt\B{ValueItem} } \\
        \hline
        \multicolumn{3}{|c|}{Value item : \B{ValueItem}} \\
        \hline
	\multicolumn{3}{c}{\B{StructItem}\alt\B{ArrayItem} } \\
        \hline
        \multicolumn{3}{|c|}{Term expression: \B{Term}} \\
        \hline
        \B{op} & : & \B{ArithOp} \\
        \B{args} & : & seq \B{Exp} \\
	\hline
        \multicolumn{3}{|c|}{Predicate: \B{Pred}} \\
        \hline
        \multicolumn{3}{c}{\B{Form} \alt \B{Comp}} \\
	\hline
        \multicolumn{3}{|c|}{Boolean formula: \B{Form}} \\
        \hline
        \B{op} & : & \B{BoolOp} \\
        \B{args} & : & seq \B{Pred} \\
\end{tabular}

\begin{tabular}[t]{rcl}
	\hline
        \multicolumn{3}{|c|}{Comparison: \B{Comp}} \\
        \hline
        \B{op} & : & \B{RelOp} \\
        \B{arg1}, \B{arg2} & : & \B{Exp} \\
        \hline
        \multicolumn{3}{|c|}{Data type: \B{Type}} \\
        \hline
        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt}\\
	\multicolumn{3}{c}{\B{StructType} \alt \B{ArrayType}}\\
	\hline
  	\multicolumn{3}{|c|}{Data struct : \B{StructType}} \\
        \hline
        \B{elements} & : & seq \B{Name} \\
	\B{type} & : & seq \B{Type}\\
	\hline
	\multicolumn{3}{|c|}{Access struct : \B{StructItem}} \\
        \hline
	\B{base} & : &  \B{Name} \\
	\B{element} & : & seq \B{Name}\\
	\B{type} & : & \B{Type}\\
	\hline
	\multicolumn{3}{|c|}{Record struct : \B{Record}} \\
        \hline
	\B{elements} & : & seq \B{Name}\\
	\B{values} & : & seq \B{Exp}\\
	\hline
        \multicolumn{3}{|c|}{Array type : \B{ArrayType}} \\
        \hline 
	\B{dom} & : & seq \B{Set}\\
	\B{ran} & : & \B{Set}\\
        \hline
        \multicolumn{3}{|c|}{Array expression : \B{Array}} \\
        \hline
	\multicolumn{3}{c}{ seq \B{Map} \alt \B{Product} \alt \B{Name}  }\\
	\hline
  	\multicolumn{3}{|c|}{Ordered pair : \B{Map}} \\
        \hline
	\B{arg1} & : & seq \B{Exp}\\ 
	\B{arg2} & : & \B{Exp}\\ 
	\hline
  	\multicolumn{3}{|c|}{Cartesian product : \B{Product}} \\
        \hline
	\B{arg1} & : & seq \B{Exp}\\ 
	\B{arg2} & : & \B{Exp}\\ 	
	\hline
	\multicolumn{3}{|c|}{Access array : \B{ArrayItem}} \\
        \hline
	\B{base} & : &  \B{Array} \\
	\B{indexes} & : & seq \B{Exp}\\
	\B{type} & : & \B{ArrayType}\\
	\hline
	\multicolumn{3}{|c|}{Simple set : \B{Set}} \\
	\hline
	\multicolumn{3}{c}{\B{BOOL} \alt \B{NAT} \alt \B{NAT$_{1}$}\alt } \\
	\multicolumn{3}{c}{\B{INT} \alt \B{Interval}} \\
	\hline
	\multicolumn{3}{|c|}{B0 interval : \B{Interval}} \\
	\hline
	\B{start}, \B{end} & : & \B{Exp} \\
	\hline
	\multicolumn{3}{|c|}{Literal : \B{Lit}} \\
	\hline
	\multicolumn{3}{c}{ \B{Bool\_lit} \alt \B{Int\_lit}} \\ 
      \end{tabular}
    \end{tabular}
  }
  }
    \caption{Abstract syntax structure of a B implementation.  For each abstract
      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
      branch in a case instruction, and the different attributes of the class),
      or a list of the possible sub-classes (e.g. \B{Inst} for the different
      kinds of instructions).
      Sequence and optional attributes are denoted ``seq'' and ``opt'',
      respectively.}
    \label{tab:node-attr}
  \end{center}
\end{figure}

\begin{figure}
  \begin{center}
    \begin{tabular}{rcl}
      \nt{module} & ::= & \ListOf{\nt{item}} \\
      \nt{item} & ::= & \nt{const\_decl} \alt \nt{function\_decl} \alt \nt{type\_def}
      \alt \nt{const\_def} \alt \nt{var\_def} \alt \nt{function\_def} \\
      \nt{const\_decl} & ::= & \nt{name} \llvm{=} \llvm{external} \llvm{constant} \nt{type} \\
      \nt{type\_def} & ::= & \nt{name} \llvm{=} \llvm{type} \nt{type} \\
      \nt{type} & ::= & \llvm{void} \alt \nt{itype} \alt \llvm{\{} \ListOf{\nt{type}} \llvm{\}} \alt \nt{type}\llvm{*} \\
      \nt{const\_def} & ::= & \nt{name} \llvm{=} \llvm{constant} \nt{type} \nt{iliteral} \\
      \nt{var\_def} & ::= & \nt{name} \llvm{=} \llvm{common} \llvm{global} \nt{type} \llvm{zeroinitializer} \\
      \nt{function\_decl} & ::= & \llvm{declare} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{type}} \llvm{)}\\
      \nt{function\_def} & ::= & \llvm{define} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{param}} \llvm{)} \llvm{\{} \ListOf{\nt{block}} \llvm{\}} \\
      \nt{param} & ::= & \nt{type} \nt{name} \\
      \nt{block} & ::= & \nt{lbl} \llvm{:} \ListOf{\nt{inst}} \\
      \nt{inst} & ::=  & \nt{name} \llvm{=} \llvm{alloca} \nt{type} \\
      & \alt & \nt{name} \llvm{=} \lalt \llvm{add} \alt \llvm{sub} \alt \llvm{mul} \alt \llvm{sdiv} \alt \llvm{srem} \ralt \nt{itype} \nt{exp} \llvm{,} \nt{exp} \\
      & \alt & \nt{name} \llvm{=} \llvm{icmp} \lalt \llvm{eq} \alt \llvm{ne} \alt \llvm{sgt} \alt \llvm{sge} \alt \llvm{slt} \alt \llvm{sle} \ralt \llvm{i1} \nt{exp} \llvm{,} \nt{exp}\\
      & \alt & \nt{name} \llvm{=} \llvm{call} \nt{type} \llvm{(} \ListOf{\nt{arg}} \llvm{)} \\
      & \alt & \nt{name} \llvm{=} \llvm{getelementptr} \nt{type} \llvm{*} 
\nt{exp}\llvm{,} \nt{index}\llvm{,} \ListOf{\nt{index}} \\
      & \alt & \nt{name} \llvm{=} \llvm{load} \nt{type} \nt{exp} \\
      & \alt & \llvm{store} \nt{type} \nt{exp}, \nt{type} \llvm{*} \nt{exp} \\
      & \alt & \nt{name} \llvm{=} \llvm{bitcast} \nt{type} \nt{name}, \nt{type} \\
      & \alt & \llvm{br} \llvm{i1} \nt{exp} \llvm{,} \llvm{label} \nt{lbl} \llvm{,} \llvm{label} \nt{lbl} \\
      & \alt & \llvm{br} \llvm{label} \nt{lbl} \\
%       & \alt & \llvm{switch} \nt{type} \nt{exp} \llvm{,} \llvm{branch} \nt{lbl} \llvm{[} \ListOf{\nt{branch}} \llvm{]} \\
      & \alt & \llvm{ret} \lalt \nt{type} \nt{exp} \alt \llvm{void} \ralt \\
      \nt{exp} & ::= & \nt{name} \alt \nt{iliteral} \alt \llvm{getelementptr} 
\llvm{(} \nt{type} \nt{exp} \llvm{,} \nt{index} \llvm{,} \ListOf{\nt{index}} \llvm{)} \\
      \nt{index} & ::= & \nt{itype} \nt{iliteral} \\
      \nt{branch} & ::= & \nt{iliteral} \nt{iliteral} \nt{lbl} \\
      \nt{arg} & ::= & \nt{type} \nt{exp}
    \end{tabular}
  \end{center}
  \caption{Grammar of the target LLVM subset: \nt{itype}, \nt{iliteral}, \nt{lbl}
    and \nt{name} correspond respectively to integer types, integer literals,
    labels and names. Choices are separated by \alt and optionally delimited by
    \lalt and \ralt.  The \ListOf{} superscript denotes a comma-separated list of
    elements of the annotated entity. }
  \label{fig:llvm-grammar}
\end{figure}

\VGM{To support arrays and structures, I added only + after index.}

@@

\paragraph{Becomes equal instructions} must evaluate the target and the source
of the assignment, and then copy the result of the latter in the former, using
the LLVM instruction \llvm{store}. The evaluation of the target is defined with
$\trad{}{lvalue}$, which yields the corresponding LLVM source code $l$ and the
variable $p$ containing the assigned location. The evaluation of the source
yields the corresponding code $r$, the variable $v$ holding the result, and its
type $t$.

@@

The expressions have elements represented by pointers that need a special 
care. The attributions between these elements are made  directly by the memory 
address.  However, the expression resolution are made by parts, where the 
partial results are stored to temporary variables. Therefore, the expression 
resolution with pointers needs an additional instruction $LOAD$ for store the 
partial results. This translation is represented by $ValueItem$.

@@

\begin{align*}
\begin{split}
  \trad{n}{\B{StructItem}} \isdef \\
   & \IF \B{n.type} \mbox{ is \B{BOOL or NAT or INT} } \THEN \\
   &\quad  \LET \ell = \lbl  \AND  \PH{t} = n.type \AND \PH{v} = n.element \IN\\
   &\quad  \llvm{"\PH{\ell} = getelementptr \%\PH{t}* @\PH{v}, i32 0, i32  \SBPos(\B{n})\nl"} \\
   &\quad  \sep  \ell  \sep \SBType(\B{n})\\
\end{split}
\end{align*}

The attribution from record structure to a structure variable, line 1 from last 
B code example, is translated by a sequence of application of rule $StructItem$.

@@

\begin{align*}
&\trad{n}{ShiftExp} \isdef \LET p \sep v \sep t  = \trad{n}{Exp} \AND \\
&\quad iStart \sep iEnd \sep iType = \trad{n}{TSymbol}  \AND  \ell = \lbl \IN \\
&\quad \IF ( iStart == 0 ) \THEN  \llvm{",\PH{t} \PH{v} "} \sep p \\
&\quad \ELSE \LET  p \sep v \sep v' \sep t'  = \trad{p,t,v,itype,istart}{Cast} \IN \\
&\quad \llvm{",\PH{t} \PH{l} "} \sep  \llvm{"\PH{p} \PH{\ell} = sub \PH{t'} \PH{v}, \PH{v'}\nl"}  \\
\end{align*}

@@

\section*{Attachments}

\subsection{Structure examples}

The following code example is defining a structure:

\begin{ccode}
MACHINE Rec
SETS
 ID={aa,bb}
VARIABLES xx, bank, total
INVARIANT
 xx : struct(name : ID, balance : NAT) & bank : struct(name : ID) &
 total:struct(money:NAT)
INITIALISATION 
 bank := rec(name:bb) ||
 xx:= rec(name:aa, balance:10) ||
 total := rec(money:20)
OPERATIONS

  res<--positive = 
  PRE xx'balance > 0 THEN
  res:=TRUE
  END;

  withrdaw(amt) = 
  PRE amt:NAT1 & (xx'balance >= amt) THEN
     xx := rec(name:xx'name, balance:(xx'balance - amt))
  END;

  unsafe_dec = 
  BEGIN
       xx'balance := xx'balance - 1
  END
END
\end{ccode}

\begin{ccode}
IMPLEMENTATION
    Rec_i

REFINES
    Rec

CONCRETE_VARIABLES
   xx ,
   bank ,
   total

INITIALISATION

   bank := rec(name:bb) ;
   xx := rec(name:aa, balance:10) ;
   total :=rec(money:20)
OPERATIONS
  res<--positive = 
  IF xx'balance > 0 THEN
       res:= TRUE 
  ELSE res := FALSE
  END;

  withrdaw(amt) = IF   amt >=0 & amt <= MAXINT & (xx'balance >= amt) THEN 
     xx := rec(name:xx'name, balance:(xx'balance - amt))
  END;

  unsafe_dec = BEGIN
       xx'balance := xx'balance -1 
  END
END
\end{ccode}

\subsection{Arrays examples}

%\iffalse  % This content is not visible to reader.
A generic example available at repository ProB\footnote{
\url{https://github.com/bendisposto/probparsers/blob/develop/bparser/src/test/resources/parsable/Array.mch}}. 
This example can be used in approach using imported components.

\subsection{A  simple example and its LLVM}

\begin{ccode}
MACHINE
   Array
VARIABLES arr

INVARIANT 
    arr : ((0..99) --> 0..1000)

INITIALISATION arr := {1|->100}

OPERATIONS

set(ix,tt)=
        PRE ix : (0..99) & tt : (0..1000)
        THEN arr := arr <+ {ix |-> tt}
        END ;

tt <-- read(ix) =
        PRE ix : (0..99)
        THEN tt := arr(ix)
        END;

swap(ix, jx)=
        PRE ix : (0..99) & jx : (0..99)
        THEN arr := arr <+ {ix |-> arr(jx), jx|-> arr(ix)}
        END

END
\end{ccode}

\begin{ccode}
IMPLEMENTATION 
   Array_i
REFINES
   Array
CONCRETE_VARIABLES
   arr
INVARIANT 
    arr :  ((0..99) --> 0..1000)
INITIALISATION
   arr(1) := 100
OPERATIONS
   set ( ix , tt ) =
   IF
       ix >= 0 & ix <=99 & tt >= 0  & tt <= 1000
   THEN

      arr(ix) := tt 
   END;

   tt <-- read ( ix ) =
   IF
       ix >= 0 & ix <=99
   THEN
       tt := arr ( ix )
   ELSE
       tt := 0
   END;

   swap ( ix , jx ) =
   IF
        ix >= 0 & ix <=99 & jx >= 0 & jx <=99 
   THEN

       VAR temp IN
           temp := arr(jx);
           arr(jx) := arr(ix);
           arr(ix) := temp
       END
   END
END
\end{ccode}

\begin{ccode}
@Array__arr = internal global [100 x i32] zeroinitializer, align 16

define void @Array__INITIALISATION() nounwind ssp {
  store i32 100, i32* getelementptr  ([100 x i32]* @Array__arr, i32 0, i64 1)
  ret void
}

define void @Array__set(i32 %ix, i32 %tt) nounwind ssp {
  %1 = alloca i32
  %2 = alloca i32
  store i32 %ix, i32* %1
  store i32 %tt, i32* %2
  %3 = load i32* %1
  %4 = icmp sge i32 %3, 0
  br i1 %4, label %5, label %19

; <label>:5                                       ; preds = %0
  %6 = load i32* %1
  %7 = icmp sle i32 %6, 99
  br i1 %7, label %8, label %19

; <label>:8                                       ; preds = %5
  %9 = load i32* %2
  %10 = icmp sge i32 %9, 0
  br i1 %10, label %11, label %19

; <label>:11                                      ; preds = %8
  %12 = load i32* %2
  %13 = icmp sle i32 %12, 1000
  br i1 %13, label %14, label %19

; <label>:14                                      ; preds = %11
  %15 = load i32* %2
  %16 = load i32* %1
  %17 = sext i32 %16 to i64
  %18 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %17
  store i32 %15, i32* %18
  br label %19
; <label>:19                                      ; preds = %14, %11, %8, %5, %0
  ret void
}
\end{ccode}

\begin{llvmcode}
define void @Array__read(i32 %ix, i32* %tt) nounwind ssp {
  %1 = alloca i32, align 4
  %2 = alloca i32*, align 8
  store i32 %ix, i32* %1, align 4
  store i32* %tt, i32** %2, align 8
  %3 = load i32* %1, align 4
  %4 = icmp sge i32 %3, 0
  br i1 %4, label %5, label %14

; <label>:5                                       ; preds = %0
  %6 = load i32* %1, align 4
  %7 = icmp sle i32 %6, 99
  br i1 %7, label %8, label %14

; <label>:8                                       ; preds = %5
  %9 = load i32* %1, align 4
  %10 = sext i32 %9 to i64
  %11 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %10
  %12 = load i32* %11
  %13 = load i32** %2, align 8
  store i32 %12, i32* %13
  br label %16

; <label>:14                                      ; preds = %5, %0
  %15 = load i32** %2, align 8
  store i32 0, i32* %15
  br label %16

; <label>:16                                      ; preds = %14, %8
  ret void
}
\end{llvmcode}

\begin{llvmcode}
define void @Array__swap(i32 %ix, i32 %jx) nounwind ssp {
  %1 = alloca i32, align 4
  %2 = alloca i32, align 4
  %temp = alloca i32, align 4
  store i32 %ix, i32* %1, align 4
  store i32 %jx, i32* %2, align 4
  %3 = load i32* %1, align 4
  %4 = icmp sge i32 %3, 0
  br i1 %4, label %5, label %30

; <label>:5                                       ; preds = %0
  %6 = load i32* %1, align 4
  %7 = icmp sle i32 %6, 99
  br i1 %7, label %8, label %30

; <label>:8                                       ; preds = %5
  %9 = load i32* %2, align 4
  %10 = icmp sge i32 %9, 0
  br i1 %10, label %11, label %30

; <label>:11                                      ; preds = %8
  %12 = load i32* %2, align 4
  %13 = icmp sle i32 %12, 99
  br i1 %13, label %14, label %30

; <label>:14                                      ; preds = %11
  %15 = load i32* %2, align 4
  %16 = sext i32 %15 to i64
  %17 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %16
  %18 = load i32* %17
  store i32 %18, i32* %temp, align 4
  %19 = load i32* %1, align 4
  %20 = sext i32 %19 to i64
  %21 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %20
  %22 = load i32* %21
  %23 = load i32* %2, align 4
  %24 = sext i32 %23 to i64
  %25 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %24
  store i32 %22, i32* %25
  %26 = load i32* %temp, align 4
  %27 = load i32* %1, align 4
  %28 = sext i32 %27 to i64
  %29 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %28
  store i32 %26, i32* %29
  br label %30

; <label>:30                                      ; preds = %14, %11, %8, %5, %0
  ret void
}

\end{llvmcode}

\subsection{Real example - Bubble Sort}

\begin{ccode}
MACHINE
   Bubble
VARIABLES
   vec,sort
INVARIANT
   vec : 0..99 --> 0..99 & sort : 0..1 &
   ( sort = 1 =>  !ii.(ii : 0..98 => vec(ii)<=vec(ii+1)))
INITIALISATION
   vec :: 0..99 --> 0..99 || sort := 0
OPERATIONS
   op_sort =
   ANY sorted_vector WHERE
    sorted_vector : 0..99 --> 0..99 &
    !ii.(ii : 0..98 => sorted_vector(ii)<=sorted_vector(ii+1))
   THEN
    vec := sorted_vector|| sort := 1
   END

END

IMPLEMENTATION
   Bubble_i
REFINES
   Bubble
CONCRETE_VARIABLES
   vec1,sort1
INVARIANT
   vec1 : 0..99 --> 0..99 & sort1 : 0..1 &
   vec = vec1 & sort1 = sort
INITIALISATION
   vec1(0) := 0  ;
   sort1 := 0
OPERATIONS
   op_sort =
   VAR nn, swapped, ii, tmp, vi, vi2 IN
	swapped := 1;
	nn := 100;
	ii:=0;
	WHILE swapped = 1 DO
		swapped := 0;
		ii:= 0;
		nn:= nn-1;

		WHILE ii<= nn DO
		vi :=  vec1(ii);
		vi2 := vec1(ii+1);
		IF vi > vi2 THEN
			tmp:= vec1(ii+1);
			vec1(ii+1):=vec1(ii);
			vec1(ii):=tmp;
			swapped:= 1
		END;
		ii:= ii+1

		INVARIANT 1>0
		VARIANT nn-ii
		END

	INVARIANT 1>0
	VARIANT nn-ii
	END
   END
END
\end{ccode}

\end{document}

>>>>

@@

\paragraph{Becomes equal instructions} \textcolor{blue}{ must evaluate the
 target and the source of the assignment, and then copy the result of the 
latter in the former, using the LLVM instruction \llvm{store}. The evaluation 
of the target is defined with $\trad{}{lvalue}$, which yields the corresponding 
LLVM source code $l$ and the variable $p$ containing the assigned location. The 
evaluation of the source yields the corresponding code $r$, the variable $v$ 
holding the result, and its type $t$. }

@@

\textcolor{blue}{ The expressions have elements represented by pointers that need a special 
care. The attributions between these elements are made  directly by the memory 
address.  However, the expression resolution are made by parts, where the 
partial results are stored to temporary variables. Therefore, the expression 
resolution with pointers needs an additional instruction $LOAD$ for store the 
partial results. This translation is represented by $ValueItem$.}

@@

\begin{align*}
\begin{split}
  \trad{n}{\B{StructItem}} \isdef \\
   & \IF \B{n.type} \mbox{ is \B{BOOL or NAT or INT} } \THEN \\
   &\quad  \LET \ell = \lbl  \AND  \PH{t} = n.type \AND \PH{v} = n.element \IN\\
   &\quad  \llvm{"\PH{\ell} = getelementptr \PH{t}* \PH{v}, i32 0, i32  \SBPos(\B{n})\nl"} \\
   &\quad  \sep  \ell  \sep \SBType(\B{n})\\
\end{split}
\end{align*}

The attribution from record structure to a structure variable, line 1 from last 
B code example, is translated by a sequence of application of rule $StructItem$.
The nested structures will be designed in the next project phase.

@@

\begin{align*}
&\trad{n}{ShiftExp} \isdef \LET p \sep v \sep t  = \trad{n}{Exp} \AND \\ \\
&\quad iStart \sep iEnd \sep iType = \trad{n}{TSymbol}  \AND  \ell = \lbl \IN \\
&\quad \IF ( iStart == 0 ) \THEN  \llvm{",\PH{t} \PH{v} "} \sep p \\
&\quad \ELSE \LET  p \sep v \sep v' \sep t'  = \trad{p,t,v,itype,istart}{Cast} \IN \\
&\quad \llvm{",\PH{t} \PH{l} "} \sep  \llvm{"\PH{p} \PH{\ell} = sub \PH{t'} \PH{v}, \PH{v'}\nl"}  \\
\end{align*}

@@
